The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-08 VegOzone.tex
Finish Section 3 - spell-check and monthly ozone analysis remain to be done.

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
As we discussed in the preceding Section~\ref{subsubsec:anomal_tpq}, meteorological conditions in 2018 were significantly different from the climatological average. How does this translate into observed ozone anomalies \chem{\Delta[O_3]}? If the data collection at Svanvik had been complete in 2018 and the climatology had recent years included, it would had been straight forward to deduct the significance of ozone anomalies in the same way as for temperature, precipitation, and global irradiance. In Section~\ref{subsubsec:clim_ozone}, we described the construction of climatologies for northern Fennoscandia based on long term observations at Esrange, Jergul/Karasjok, and Pallas. For Svanvik this assessment is based on rather uncertain observations from 1986--1996. We also discussed the general patterns, similarities and differences. We shall now investigate \chem{\Delta[O_3]} in 2018/19 with respect to these climatologies in more detail.

In the following, we will use the interchangable terms anomalies and residuals to distinquish between a climatological (anomalies) and a statistical (residuals) point of view.
All the above factors make it neccessary to look not only at anomalies but statistically assess ozone concentration residuals at Svanvik as well as at Esrange and Pallas. Since the climatology of latter is more reliable, we will start with surface ozone concentration residuals \chem{\Delta[O_3]} at Esrange and Pallas in 2018. As data for 2019 has not yet been available from EBAS (last access, February 2020), we refrain from looking into these.

@@

Figure~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik} displays the probability density functions for Svanvik based on data taken in 2018/19 as well as relative to the climatology derived for northern Fennoscandia (denoted with NF in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}a) and the climatology derived for Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}b).
As previously mentioned, \chem{\left<[O_3]\right>} at Svanvik are lower compared to the northern Fennoscandic climatology ($(6.6\pm 7.1)\,\unit{ppb}$). For 2018 and 2019, we find $\Delta\chem{\left<[O_3]\right>} = (-3.8\pm 5.6)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (-5.2\pm 5.0)\,\unit{ppb}$, respectively. While the probablility density of the residuals between the two climatologies is skewed towards negative values, the skew of 2018/19 is positive.
With respect to the climatology for Svanvik, the probability density functions average for 2018/19 lies at $\Delta\chem{\left<[O_3]\right>} = (2.6\pm 5.8)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (1.2\pm 5.0)\,\unit{ppb}$, respectively.
Neither of the \chem{\Delta\left<[O_3]\right>} is significant by statistical means. But daily average ozone concentrations are slightly elevated in 2018 compared to 2019.

@@

To further investigate this, we have a look at a Student's~t-test ($t = \frac{\chem{\Delta[O_3]}}{\sigma}$), assuming the same sample uncertainty $\sigma$ in both, climatology and the year in question. Results for Esrange and Pallas are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test}. This t-test clearly shows, that at both sites for at least $50\,\unit{\%}$ of days the deviation from climatological mean is significant on the $1\sigma$ level. In Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}, the test statistics for Svanvik are displayed. From Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}a), we can deduce, that the deviation from the northern Fennoscandic climatology is significant on the $2\sigma$ level for at least half of the days in the Svanvik climatology, while the deviation is less certain in 2018 and 2019. In Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}b) the same is shown with respect to the climatology from Svanvik. We find that in less than $50\,\unit{\%}$ of days the ozone concentration deviation is significant on the $1\sigma$ level.

>>>>

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
As we discussed in the preceding Section~\ref{subsubsec:anomal_tpq}, meteorological conditions in 2018 were significantly different from the climatological average. How does this translate into observed ozone anomalies \chem{\Delta[O_3]}?

If the data collected at Svanvik had been complete in 2018 and the climatology had extended into the resent past, we could have assessed the significance of ozone anomalies in the same way as for temperature, precipitation, and global irradiance. In Section~\ref{subsubsec:clim_ozone}, we described the construction of climatologies for northern Fennoscandia based on long term observations at Esrange, Jergul/Karasjok, and Pallas. For Svanvik this assessment is based on rather uncertain observations from 1986--1996. We also discussed the general patterns, similarities and differences. We shall now investigate \chem{\Delta[O_3]} in 2018/19 with respect to these climatologies in more detail.

In the following, we will use the interchangable terms anomalies and residuals to distinquish between a climatological (anomalies) and a statistical (residuals) point of view.
All the above factors make it neccessary to look not only at anomalies but statistically assess ozone concentration residuals at Svanvik as well as at Esrange and Pallas in comparison. Since the climatology of latter is more reliable, we will start with surface ozone concentration residuals \chem{\Delta[O_3]} at Esrange and Pallas in 2018. As data for 2019 has not yet been available from EBAS (last access, February 2020), we refrain from looking into these.

@@

Figure~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik} displays the probability density functions of daily mean ozone concentration residuals for Svanvik based on data taken in 2018/19 with respect to the climatology derived for northern Fennoscandia (denoted NF in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}a) and the climatology derived for Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}b).
As previously mentioned, \chem{\left<[O_3]\right>} at Svanvik are $(6.6\pm 7.1)\,\unit{ppb}$ lower than the northern Fennoscandic climatology. For 2018 and 2019, we find $\Delta\chem{\left<[O_3]\right>} = (-3.8\pm 5.6)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (-5.2\pm 5.0)\,\unit{ppb}$, respectively. While the probablility density of the residuals between the two climatologies is skewed towards negative values, the skew of 2018/19 is positive.
With respect to the climatology for Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}b), the probability density functions average for 2018/19 lies at $\Delta\chem{\left<[O_3]\right>} = (2.6\pm 5.8)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (1.2\pm 5.0)\,\unit{ppb}$, respectively.
Neither of the \chem{\Delta\left<[O_3]\right>} is significant by statistical means, but daily average ozone concentrations are slightly elevated in 2018 compared to 2019.

The \chem{\left<\Delta\left<[O_3]\right>\right>} derived for Svanvik in 2018 is slightly larger than for Esrange and Pallas. This could mean either that 2018 had been more exceptional at Svanvik or the Svanvik climatology may need a correction for increased tropospheric background ozone concentration. Looking at daily \chem{\Delta_{present-past}\left<[O_3]\right>}, wherein present is defined as the average of 2010--2018 and past as the respective beginning of data taking until 1997, does not reveal a clear trend: We find for Esrange $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas $(-2.5\pm 3.7)\,\unit{ppb}$.

@@

To further investigate the ozone residuals, we conduct a Student's~t-test ($t = \frac{\chem{\Delta[O_3]}}{\sigma}$), assuming the same sample uncertainty $\sigma$ in both, climatology and the year in question. Results for Esrange and Pallas are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test}. This t-test clearly shows, that at both sites for at least $50\,\unit{\%}$ of days the residuals are significant on the $1\sigma$ level. In Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}, the test statistics for Svanvik are displayed. From Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}a), we can deduce, that the residuals of the Svanvik climatology from the northern Fennoscandic climatology are significant on the $2\sigma$ level for at least half of the days, while the residuals are less certain in 2018 and 2019. In Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}b) the same is shown for the years 2018/19 with respect to the climatology from Svanvik. We find that in less than $50\,\unit{\%}$ of days the ozone concentration residuals is significant on the $1\sigma$ level.
